```{r quickstart-setup, include=FALSE, purl=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "quick-")
```

# Quick Start  {#chap-quick-start}

The default MARSS model is:
\begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}_t\mathbf{x}_{t-1} + \mathbf{u}_t + \mathbf{C}_t\mathbf{c}_t + \mathbf{G}_t\mathbf{w}_t, \text{ } \mathbf{w}_t \sim \,\text{MVN}(0,\mathbf{Q}_t) \\
\mathbf{y}_t = \mathbf{Z}_t\mathbf{x}_t + \mathbf{a}_t + \mathbf{D}_t\mathbf{d}_t + \mathbf{H}_t\mathbf{v}_t, \text{ } \mathbf{v}_t \sim \,\text{MVN}(0,\mathbf{R}_t) \\
\mathbf{x}_1 \sim \,\text{MVN}(\boldsymbol{\pi},\boldsymbol{\Lambda}) \text{ or } \mathbf{x}_0 \sim \,\text{MVN}(\boldsymbol{\pi},\boldsymbol{\Lambda})
\end{gathered}
(\#eq:marss-quick)
\end{equation}
$\mathbf{c}$ and $\mathbf{d}$ are inputs (aka, exogenous variables or covariates or indicator variables) and must have no missing values. In most cases, $\mathbf{G}$ and $\mathbf{H}$ are fixed (not estimated) and must have no missing values. $\mathbf{w}_t$ and $\mathbf{v}_t$ are uncorrelated. The $\mathbf{R}$, $\mathbf{Q}$ and $\boldsymbol{\Lambda}$ variance-covariance matrices can be set to zero (or partially to zero) to specify partially deterministic systems.

If you already work with models in the form of Equation \@ref(eq:marss-quick), you can immediately fit your model with the **MARSS** package.  Install the **MARSS** package and then type `library(MARSS)` at the command line to load the package.

The call to fit a model in **MARSS** is 
```
fit <- MARSS(data, model=model.list)
```
The data must be entered as a $n \times T$ matrix, a `ts` object (which will be converted to a $n \times T$ matrix) or a vector (which will be converted to a $1 \times T$ matrix). No missing time steps are allowed. Data missing for a time step are specified with NA. NAs may appear anywhere in the data and do not need to line up across the rows in `data`.

`model.list` is a list that specifies the form of each of the parameter matrices in Equation \@ref(eq:marss-quick). See next section and Chapter \@ref(chap-model-list).

There are plot (and autoplot), print, summary, coef, fitted, tsSmooth, residuals and predict (and forecast) functions for marssMLE objects. See the man files, e.g. `?plot`, and Part \@ref(part3) for examples.


## Model specification

Model specification for `MARSS()` is one-to-one to your MARSS model written in matrix form. You can mix fixed and estimated parameters, linear constraints, and time-varying parameters. Zeros are allowed on the diagonals of $\mathbf{Q}$, $\mathbf{R}$ and $\boldsymbol{\Lambda}$ to specify partially deterministic systems. There are text shortcuts for common forms of the parameter matrices and there are default forms. See Chapter \@ref(chap-model-list).

Example: a mean-reverting random walk model with three observation time series:
\begin{gather*}
\begin{bmatrix}x_1\\ x_2\end{bmatrix}_t
= \begin{bmatrix}b&0\\ 0&b\end{bmatrix}
\begin{bmatrix}x_1\\ x_2\end{bmatrix}_{t-1}
+ \begin{bmatrix}w_1\\ w_2\end{bmatrix}_t, \quad 
\begin{bmatrix}w_1\\ w_2\end{bmatrix}_t \sim \,\text{MVN}\begin{pmatrix}\begin{bmatrix}0\\0\end{bmatrix},\begin{bmatrix}q_{11}&q_{12}\\ q_{12}&q_{22}\end{bmatrix} \end{pmatrix}
\\
\begin{bmatrix}y_1\\y_2\\y_3\end{bmatrix}_t
= \begin{bmatrix}1&1\\ 0&1\\ 1&0\end{bmatrix}
\begin{bmatrix}x_1\\x_2\end{bmatrix}_t
+ \begin{bmatrix}v_1\\ v_2\\ v_3\end{bmatrix}_t,\quad 
\begin{bmatrix}v_1\\ v_2\\ v_3\end{bmatrix}_t \sim MVN\begin{pmatrix}\begin{bmatrix}a_1\\ 0\\ 0\end{bmatrix},
 \begin{bmatrix}r_{11}&0&0\\ 0&r&0\\ 0&0&r\end{bmatrix} \end{pmatrix} \\
\begin{bmatrix}x_1\\ x_2\end{bmatrix}_0 \sim \,\text{MVN}\begin{pmatrix}\begin{bmatrix}0\\ 0\end{bmatrix},\begin{bmatrix}1&0\\ 0&1\end{bmatrix} \end{pmatrix} \\
\end{gather*}

To fit with `MARSS()`, we translate this model into equivalent matrices (or arrays if time-varying) . Matrices that combine fixed and estimated values are specified using a list matrix with numerical values for fixed values and  character names for the estimated values. 
```{r quick-start-01}
B1 <- matrix(list("b", 0, 0, "b"), 2, 2)
U1 <- matrix(0, 2, 1)
Q1 <- matrix(c("q11", "q12", "q12", "q22"), 2, 2)
Z1 <- matrix(c(1, 0, 1, 1, 1, 0), 3, 2)
A1 <- matrix(list("a1", 0, 0), 3, 1)
R1 <- matrix(list("r11", 0, 0, 0, "r", 0, 0, 0, "r"), 3, 3)
pi1 <- matrix(0, 2, 1)
V1 <- diag(1, 2)
model.list <- list(B = B1, U = U1, Q = Q1, 
                   Z = Z1, A = A1, R = R1, 
                   x0 = pi1, V0 = V1, tinitx = 0)
```

If you print these out, you will see the one-to-one correspondence between the model in R and the math version of the model.  Matrix names in the model list must be `B`, `U`, `C`, `c`, `Q`, `Z`, `A`, `D`, `d`, `R`, `x0`, and `V0`. In many cases you will use the default values and do not need to specify the parameter for the model list.  The `tinitx` specifies whether the initial state for $x$ is at $t=1$ (`tinitx=1`) or $t=0$ (`tinitx=0`).  

The **MARSS** package is designed to handle linear constraints within the parameter matrices.  Linear constraint means you can write the elements of the matrix as a linear equation of all the other elements. Time-varying parameters can also be handled.

## Linear constraints

Your model can have simple linear constraints within all the parameters except $\mathbf{Q}$, $\mathbf{R}$ and $\boldsymbol{\Lambda}$.  For example $1+2a-3b$ is a linear constraint. When entering this value for you matrix, you specify this as `"1+2*a+-3*b"`. NOTE: `+`'s join parts so use `+-3*b` to specify $-3b$. Anything after `*` is a parameter. So `1*1` has a parameter called `"1"`. Example, let's change the matrices in the previous model to:
\begin{equation*}
\mathbf{B} = \begin{bmatrix}b-0.1&0\\ 0&b+0.1\end{bmatrix}\quad
\mathbf{Q} = \begin{bmatrix}q_{11}&0\\ 1&0\end{bmatrix}\quad
\mathbf{Z} = \begin{bmatrix}z_1-z_2&2 z_1\\ 0&z_1\\ z_2&0\end{bmatrix}
\end{equation*}

This would be specified as (notice `"1*z1+-1*z2"` for `z1-z2`):
```{r}
B1 <- matrix(list("-0.1+1*b", 0, 0, "0.1+1*b"), 2, 2)
Q1 <- matrix(list("q11", 0, 0, 1), 2, 2)
Z1 <- matrix(list("1*z1+-1*z2", 0, "z2", "2*z1", "z1", 0), 3, 2)
model.list <- list(B = B1, U = U1, Q = Q1, Z = Z1, A = A1, R = R1, x0 = pi1, V0 = V1, tinitx = 0)
```

You can call `toLatex()` on your model to make sure you and `MARSS()` agree on what model you are trying to fit:
```
fit <- MARSS(data, model=model.list)
toLatex(fit$model)
```

## Time-varying parameters

You can pass in an array of $T$ matrices for a time-varying parameter. $T$ is the number of time-steps in your data and is the 3rd dimension in the parameter array.  
\begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}_t\mathbf{x}_{t-1} + \mathbf{u}_t + \mathbf{C}_t\mathbf{c}_t + \mathbf{G}_t\mathbf{w}_t, \quad
\mathbf{W}_t \sim \,\text{MVN}(0,\mathbf{Q}_t)\\
\mathbf{y}_t = \mathbf{Z}_t\mathbf{x}_t + \mathbf{a}_t + \mathbf{D}_t\mathbf{d}_t + \mathbf{H}_t\mathbf{v}_t, \quad
\mathbf{V}_t \sim \,\text{MVN}(0,\mathbf{R}_t)\\
\mathbf{x}_{t_0} \sim \,\text{MVN}(\boldsymbol{\pi},\boldsymbol{\Lambda})
\end{gathered}
\end{equation}
Note the time indexing.  Make sure you enter your arrays such that the right parameter (or input) at time $t$ lines up with $\mathbf{x}_t$, e.g. it is common for state equations to have $\mathbf{B}_{t-1}$ lined up with $\mathbf{x}_t$ so you might need to enter the $\mathbf{B}$ array such that your $\mathbf{B}_{t-1}$ is entered at `Bt[,,t]` in the R code.

The length of the 3rd dimension must be the same as your data.  For example, say in your  mean-reverting random walk model (the example on the first page) you wanted $\mathbf{B}(2,2)$ to be one value before $t=20$ and another value after but $\mathbf{B}(1,1)$ to be time constant. You can pass in the following:
```
TT <- dim(data)[2]
B1 <- array(list(),dim=c(2,2,TT))
B1[,,1:20] <- matrix(list("b",0,0,"b_1"),2,2)
B1[,,21:TT] <- matrix(list("b",0,0,"b_2"),2,2)
```
Notice the specification is one-to-one to your $\mathbf{B}_t$ matrices on paper.

## Inputs `c` and `d`

In the MARSS model,  $\mathbf{C}$ and $\mathbf{D}$ are the estimated parameters and $\mathbf{c}$ and $\mathbf{d}$ are the inputs.  In the model list, `c` and `d` are entered as a matrix with $T$ columns. The number of rows depends on the number of inputs.

Let's say you have temperature data and you want to include a linear effect of temperature that is different for each $\mathbf{x}$ time series:
```
C1 <- matrix(c("temp1","temp2"),2,1)
model.list <- list(B=B1, U=U1, C=C1, c=temp, Q=Q1,
                   Z=Z1, A=A1, R=R1, x0=pi1,V 0=V1, tinitx=0)
```
If you want a factor effect, then you will need to recode your factor as a matrix with $T$ columns and a row for each factor.  See Chapter \@ref(chap-covariates) for an example.

## Tips and Tricks

* Use `plot(fit)` (or `autoplot(fit)`) to see a series of plots and diagnostics for your model. 
* Try `MARSSinfo()` if you get errors you don't understand or fitting is taking a long time to converge.  
* When fitting a model with `MARSS()`, pass in `silent=2` to see what `MARSS()` is doing.  This puts it in verbose mode.   
* Use `fit=FALSE` to set up a model without fitting. Let's say you do `fit <- MARSS(..., fit=FALSE)`.  Now you can do `summary(fit$model)` to see what `MARSS()` thinks you are trying to fit.  You can also try `toLatex(fit$model)` to make a LaTeX file and pdf version of your model (saved in the working directory).  This loads the **Hmisc** package (and all its dependencies) and requires that you are able to process LaTeX files (e.g. you have the **tinytex** package).
* Let's say you specified your model with some text shortcuts, like `Q="unconstrained"`, but you want the list matrix form for a next step. `a <- summary(fit$model)` returns that list (invisibly).  Because the model argument of `MARSS()` will understand a list of list matrices, you can pass in `model=a` to specify the model.  
* `MARSSkfas(fit, return.kfas.model=TRUE)`  will return your model in **KFAS** form (class SSModel), thus you can use all the functions available in the **KFAS** package on your model.
