```{r examples-output-setup, include=FALSE, purl=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "quick-output-")
set.seed(123)
```
# Common output for fits {#chap-short-output}

This is a brief summary of common outputs for model fits. The output are shown with the following simple fit:
```{r quick-unis-output-dat, results="hide"}
dat <- cumsum(rnorm(100,0,0.5)) + rnorm(100,0,0.5)
fit <- MARSS(dat)
```

Use the **broom** package to get the model output in **tidy** form. The confidence intervals shown are approximate and based on the estimated Hessian matrix. See `?tidy.marssMLE` for information on how to change to a different type of confidence interval.

```{r quick-tidy}
broom::tidy(fit)
```

The get the estimated states use:
```{r quick-augment1}
head(stats::tsSmooth(fit, type="xtT"))
```
Note that generic rownames were given since none were specified for `dat`.
You can also get all the Kalman filter and smoother estimates for $\mathbf{x}$ from `MARSSkf()`. See `?MARSSkf`.

There are two types of fitted values that are used in the state-space literature: the one-step-ahead which uses on the data up to $t-1$ and the smoothed fitted values which uses all the data. Read up on fitted values for MARSS models at `?fitted.marssMLE`.

To get the fitted values, the estimated $\mathbf{y}$, using all the data:
```{r quick-ytT}
head(fitted(fit, type="ytT"))
```
To get the fitted values, the estimated $\mathbf{y}$, using the data up to time $t-1$:
```{r quick-ytt1}
head(fitted(fit, type="ytt1"))
```
Note, this is the default for `fitted(fit)`.

To make some generic plots and diagnostic plots, you can use `autoplot()` in the **ggplot2** package:
```{r quick-autoplot}
ggplot2::autoplot(fit, plot.type="xtT")
```

To get residuals, use:
```{r quick-resids}
resids <- residuals(fit)
head(format(resids, digits=2))
```
There are many types of residuals possible for MARSS models. The default `residuals(fit)` will return the innovations residuals, which are what are typically used for residuals diagnostics for state-space models. However `residuals()` and the companion `MARSSresiduals()` will return all possible residuals for your model. Read about MARSS residuals in `?residuals.marssMLE`.

To get your parameter estimates, use:
```{r quick-coef, results="hide"}
coef(fit)
```
Type `?coef.marssMLE` to see the different forms that you can get your estimated parameters in. A common form is as a matrix. To get that use `coef(fit, type="matrix")`.
```{r quick-coef2}
coef(fit, type="matrix")$Q
```


