# <i style="font-size: 110%; color:#990017;">Part 2. Examples</i> {- #part2}

Here a series of short examples are shown for different types of models specified in MARSS structure. Output is shown briefly.

The MARSS model has the form:
\begin{equation}(\#eq:marss-part2)
\begin{gathered}
\mathbf{x}_t = \mathbf{B}_t\mathbf{x}_{t-1} + \mathbf{u}_t + \mathbf{C}_t\mathbf{c}_t + \mathbf{G}_t\mathbf{w}_t, \text{ where } \mathbf{w}_t \sim \,\text{MVN}(0,\mathbf{Q}_t)\\
\mathbf{y}_t = \mathbf{Z}_t\mathbf{x}_t + \mathbf{a}_t + \mathbf{D}_t\mathbf{d}_t + \mathbf{H}_t\mathbf{v}_t, \text{ where } \mathbf{v}_t \sim \,\text{MVN}(0,\mathbf{R}_t)\\
\mathbf{x}_1 \sim \,\text{MVN}(\boldsymbol{\pi},\boldsymbol{\Lambda}) \text{ or } \mathbf{x}_0 \sim \,\text{MVN}(\boldsymbol{\pi},\boldsymbol{\Lambda})
\end{gathered}
\end{equation} 

The $\mathbf{x}$ on the left are the hidden states. The $\mathbf{y}$ on the left are the observed data. Missing values are allowed in $\mathbf{y}$. The $\mathbf{c}$ and $\mathbf{d}$ are inputs (not estimated).  Bolded capitalized values on the right are parameters and 2D (or 3D if time-varying) matrices. Parameters can be estimated, constrained or fixed at a specific value. Within a parameter matrix, you can have a combination of estimated, constrained (shared), or fixed values. The $\mathbf{u}$ and $\mathbf{a}$ are parameter column-matrices and can be similarly estimated, constrained or fixed. $\mathbf{w}$ and $\mathbf{v}$ are the errors and are computed values, after parameters are estimated.

### Output

This is a brief summary of outputs. The output are shown with the following simple fit:
```{r quick-unis-output-dat, results="hide"}
dat <- cumsum(rnorm(100,0,0.5)) + rnorm(100,0,0.5)
fit <- MARSS(dat)
```

Use the **broom** package to get the model output in **tidy** form. The confidence intervals shown are approximate and based on the estimated Hessian matrix. See `?tidy.marssMLE` for information on how to change to a different type of confidence interval.

```{r quick-tidy}
broom::tidy(fit)
```

The get the estimated states use:
```{r quick-augment1}
head(stats::tsSmooth(fit, type="xtT"))
```
Note that generic rownames were given since none were specified for `dat`.
You can also get all the Kalman filter and smoother estimates for $x$ from `MARSSkf()`. See `?MARSSkf`.

There are two types of fitted values that are used in the state-space literature: the one-step-ahead which uses on the data up to $t-1$ and the smoothed fitted values which uses all the data. Read up on fitted values for MARSS models at `?fitted.marssMLE`.

To get the fitted values, the estimated $y$, using all the data:
```{r quick-ytT}
head(fitted(fit, type="ytT"))
```
To get the fitted values, the estimated $y$, using the data up to time $t-1$:
```{r quick-ytt1}
head(fitted(fit, type="ytt1"))
```
Note, this is the default for `fitted(fit)`.

To make some generic plots and diagnostic plots, you can use `autoplot()` in the **ggplot2** package:
```{r quick-autoplot}
ggplot2::autoplot(fit, plot.type="xtT")
```

To get residuals, use:
```{r quick-resids}
residuals(fit)
```
There are many types of residuals possible for MARSS models. The default `residuals(fit)` will return the innovations residuals, which are what are typically used for residuals diagnostics for state-space models. However `residuals()` and the companion `MARSSresiduals()` will return all possible residuals for your model. Read about MARSS residuals in `?residuals.marssMLE`.

To get your parameter estimates, use:
```{r quick-coef}
coef(fit)
```
Type `?coef.marssMLE` to see the different forms that you can get your estimated parameters in. A common form is as a matrix. To get that use ```coef(fit, type="matrix")```.


