```{r covariates-setup, include=FALSE, purl=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "input-")
```

# Covariates format {#chap-covariates}

Covariates (aka inputs or exogenous variables) appear as $\mathbf{c}$ and $\mathbf{d}$ in the MARSS model and their effects are in the $\mathbf{C}$ and $\mathbf{D}$ matrices. As usual, constraints (such as shared or fixed values) can be applied within the matrix, though not across the $\mathbf{C}$ and $\mathbf{D}$ matrices. No missing values are allowed in $\mathbf{c}$ and $\mathbf{d}$.

\begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}_t\mathbf{x}_{t-1} + \mathbf{u}_t + \mathbf{C}_t\mathbf{c}_t + \mathbf{G}_t\mathbf{w}_t, \text{ } \mathbf{w}_t \sim \,\text{MVN}(0,\mathbf{Q}_t) \\
\mathbf{y}_t = \mathbf{Z}_t\mathbf{x}_t + \mathbf{a}_t + \mathbf{D}_t\mathbf{d}_t + \mathbf{H}_t\mathbf{v}_t, \text{ } \mathbf{v}_t \sim \,\text{MVN}(0,\mathbf{R}_t) \\
\mathbf{x}_1 \sim \,\text{MVN}(\boldsymbol{\pi},\boldsymbol{\Lambda}) \text{ or } \mathbf{x}_0 \sim \,\text{MVN}(\boldsymbol{\pi},\boldsymbol{\Lambda})
\end{gathered}
(\#eq:marss-cov)
\end{equation}

Covariates are passed in the model list.
```
MARSS(data, model=list(c=..., d=...))
```
Your covariates must a matrix with time across the columns ($p \times T$ matrix) where $p$ is the number of covariates. No NAs are allowed in covariates. See the chapters in the `r MARMES` book dealing with missing values in your covariates.

Here is an example of a covariate matrix for 2 covariates.
\begin{equation*}
\mathbf{y} = \left[ \begin{array}{ccccccc}
    1 & 2 & 5 & 7 & 3.2 & 8\\
    2 & 5 & 3 & 8 & 5.1 & 10.2\end{array} \right]
\end{equation*}


### Factor covariates {-}

If your covariates are factors, like site number or month, and you are estimating a $a$ value, i.e. level or intercept, for each then you will use a matrix with 0s and 1s. This is identical to how say `lm()` would translate your model with factors.

Say your covariate is quarters and you have 3 years of data:
```{r}
covariate <- rep(paste0("q",1:4), 3)
```
You translate this to a matrix with four rows and 10 columns. Each row is for a different quarter.
```{r}
vals <- unique(covariate)
TT <- length(covariate)
p <- length(vals)
c <- matrix(0, p, TT)
for(i in 1:p) c[i,] <- covariate == vals[i]
rownames(c) <- vals
c
```

