```{r covariates-setup, include=FALSE, purl=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "input-")
```

# Covariates format {#chap-covariates}

The first argument to `MARSS()` is your data, aka your response variables or the $\mathbf{y}$ in your MARSS equation.
```
MARSS(data, model=list(c=..., d=...))
```

Your covariates must a matrix with time across the columns ($p \times T$ matrix) where $p$ is the number of covariates. No NAs are allowed in covariates. See [MARMES](https://nwfsc-timeseries.github.io/MARMES/) chapters dealing with missing values in your covariates.

Here is an example of a covariate matrix for 2 covariates.
\begin{equation*}
\mathbf{y} = \left[ \begin{array}{ccccccc}
    1 & 2 & 5 & 7 & 3.2 & 8\\
    2 & 5 & 3 & 8 & 5.1 & 10.2\end{array} \right]
\end{equation*}


## Factor covariates

If your covariates are factors, like site number or month, and you are estimating a $a$ value, i.e. level or intercept, for each then you will use a matrix with 0s and 1s. This is identical to how say `lm()` would translate your model with factors.

Say your covariate is quarters and you have 3 years of data:
```{r}
covariate <- rep(paste0("q",1:4), 3)
```
You translate this to a matrix with four rows and 10 columns. Each row is for a different quarter.
```{r}
vals <- unique(covariate)
TT <- length(covariate)
p <- length(vals)
c <- matrix(0, p, TT)
for(i in 1:p) c[i,] <- covariate == vals[i]
rownames(c) <- vals
c
```

