```{r covariates-setup, include=FALSE, purl=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "input-")
```

# Covariates format {#chap-covariates}

Covariates (aka inputs or exogenous variables) appear as $\cc$ and $\dd$ in the MARSS model and their effects are in the $\CC$ and $\DD$ matrices. As usual, constraints (such as shared or fixed values) can be applied within the matrix, though not across the $\CC$ and $\DD$ matrices. No missing values are allowed in $\cc$ and $\dd$.

\begin{equation}
\begin{gathered}
\xx_t = \BB_t\xx_{t-1} + \uu_t + \CC_t\cc_t + \GG_t\ww_t, \text{ } \ww_t \sim \MVN(0,\QQ_t) \\
\yy_t = \ZZ_t\xx_t + \aa_t + \DD_t\dd_t + \HH_t\vv_t, \text{ } \vv_t \sim \MVN(0,\RR_t) \\
\xx_1 \sim \MVN(\pipi,\LAM) \text{ or } \xx_0 \sim \MVN(\pipi,\LAM)
\end{gathered}
(\#eq:marss-cov)
\end{equation}

Covariates are passed in the model list.
```
MARSS(data, model=list(c=..., d=...))
```
Your covariates must a matrix with time across the columns ($p \times T$ matrix) where $p$ is the number of covariates. No NAs are allowed in covariates. See the chapters in the `r MARMES` book dealing with missing values in your covariates.

Here is an example of a covariate matrix for 2 covariates.
\begin{equation*}
\yy = \left[ \begin{array}{ccccccc}
    1 & 2 & 5 & 7 & 3.2 & 8\\
    2 & 5 & 3 & 8 & 5.1 & 10.2\end{array} \right]
\end{equation*}


### Factor covariates {-}

If your covariates are factors, like site number or month, and you are estimating a $a$ value, i.e. level or intercept, for each then you will use a matrix with 0s and 1s. This is identical to how say `lm()` would translate your model with factors.

Say your covariate is quarters and you have 3 years of data:
```{r}
covariate <- rep(paste0("q",1:4), 3)
```
You translate this to a matrix with four rows and 10 columns. Each row is for a different quarter.
```{r}
vals <- unique(covariate)
TT <- length(covariate)
p <- length(vals)
c <- matrix(0, p, TT)
for(i in 1:p) c[i,] <- covariate == vals[i]
rownames(c) <- vals
c
```

